A formal approach to child survival from the perspective of mothers

### 2019-08-07 draft.Rmd
first draft of dag revisions 1

---
title: "A formal approach to child survival from the perspective of mothers"
author: "DA-IW"
date: "August 2019"
output: pdf_document
bibliography: bib.bibtex
---

@@

An important relation in kinship theory is the expected descendants for mothers at certain age. Child survival and child death affects important dimensions of a mothers' life, such as their reproduction and quality of life.

@@

Existing formal demographic tools are used within a stable population framework to characterize child mortality from the perspective of mothers considering both an absolute and a relative change in the force of mortality.

@@

The expected change in child survival given an absolute change in mortality depends on the difference between maternal age and the age distribution o her surviving children. For absolute changes in mortality, it is necessary to consider a measure of entropy weighted by fertility.

@@

The paper proposes related expressions to describe how mortality changes affects child survival for mothers across demographic regimes. The paper also proposes alternative ways of conceptualizing child survival.

## Relationship

@@

\begin{equation}
\label{eq:CS}
CS_a = \int_{\alpha}^a{m_{x}l_{a-x}dx}
\end{equation}

@@

In equation \ref{eq:abs_change}, the expected change in $CS_a$ is inversely proportional to the difference between maternal age \textit{a} and the mean age of the mother at the birth of her surviving daughters $\overline{x}_{(CS)}$. In other words, the magnitude of the change depends negatively on the age distribution of the surviving offspring (younger children experience longer periods of exposure to risk).

@@

@@

But maybe a mother would relate this lost years with the expected ones, feeling the *intensity* of her losts (*Intensity Time Lost*), and make possible the comparison between fertility regimes:

## Proofs

### Absolute change

Considering that $m_{x,\delta}=m_{x}+\delta$, $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$, and:

@@

### Relative change

@@

Using that the derivative of $dCS_{(a)}^\delta / d\delta = log(l_{a-x}) l_{a-x}^{(1+\delta)}$, evaluating $\delta$ near zero we get the final expression:

@@

@@

### Mean Age at Lost

Start with the integral of the age $x+t$ at each death child weighted by fertility and survival function

@@

## History

Equation  \ref{eq:CS} was developed in @goodman1974, in the "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$).

@lotka1931orphanhood started to model the orphanhood in a theorical population, comparing the effect in different regimes.  The expected number of living daughters as a stable population concept was mentioned in @goodman1974, extending this idea to the rest of the kinship, in various genealogical directions. The formal effect of different kind of changes in mortality by age was well described in @wrycza2012.

## Applications

The proposed indexes will be applied to two very different regimes.
The changes of mortality will be done computaionally to evaluate which component is more important in the raltive case

## Bibliography

>>>>

---
title: "A formal approach to child survival from the perspective of mothers"
author: "Ivan Williams and Diego Alburez-Gutierrez"
date: "August 2019"
output: pdf_document
bibliography: bib.bibtex
---

@@

The expected number of descendants for mothers that reach a certain age is an important relation in kinship theory. Child survival (and child death) affect mothers in a variety of ways, yet it is not clear how changes in mortality affect the number of surviving children for mothers.

@@

Existing formal demographic tools are used within a stable population framework to characterize child survival from the point of view of a mother, considering both absolute and relative changes in mortality patterns.

@@

The expected change in child survival given an absolute change in mortality depends on the difference between maternal age and the age structure of her surviving children. The effects of absolute changes in mortality are explained using a fertility-weighted measure of entropy.

@@

The paper proposes new expresions to characterize the effects of changes in mortality on child survival. The different meassures it proposes can be applied to populations at different stages of the demographic transition.

## 1. Relationship

@@

\begin{equation}
\label{eq:CS}
CS_a = \int_{\alpha}^a{m_{x}l_{a-x}dx}.
\end{equation}

@@

In this equation, the expected change in child survival is inversely proportional to the difference between maternal age \textit{a} and the mean age of the mother at the birth of her surviving daughters $\overline{x}_{(CS)}$. In other words, the magnitude of the change depends negatively on the age distribution of the surviving offspring (as younger children experience longer periods of exposure to risk).

@@

<!-- Aqui no cambie nada porque no entiendo muy bien lo que hiciste. Talvez hablamos para discutir? -->

@@

## 2. Proofs

### 2.1. Absolute change

Considering that $m_{x,\delta}=m_{x}+\delta$ and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$:

@@

### 2.2. Relative change

@@

Using the derivative of $dCS_{(a)}^\delta / d\delta = log(l_{a-x}) l_{a-x}^{(1+\delta)}$ and evaluating $\delta$ near zero we get the final expression:

@@

<!-- I think this needs re-weriting for clarity but I didn't want to change it to avoid messing it up! -->

<!-- Podrias dar alguna referencia para lo de entropia?  -->

@@

## 3. History

Kinship theory has a long pedigree in mathematical demography. @lotka1931orphanhood started to model the orphanhood in a theorical population, comparing the effect in different demographic regimes. Equation  \ref{eq:CS} was originally proposed by @brass_derivation_1953 to estimate child mortality. The notion of estimating the expected number of living daughters in a stable population was extended by @goodman1974 to other kin relations (granddaughters, cousins, etc.). The "counting method" approach, being the net reproduction rate $R_0$ limited to $a$, or the (not complete) ratio between generations (multiplying and dividing by the mother generation $B$), was further popularized in @KeyfitzMath. @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'. More recently, @wrycza2012 has looked at the formal effect of different kind of changes in mortality by age in a paper published in this series.

## 4. Related meassures

### 4.1. Time with children lost

But maybe a mother would relate these lost years with the expected ones, feeling the *intensity* of her losts (*Intensity Time Lost*), and make possible the comparison between fertility regimes:

### 4.2. Mean Age at Child Loss

Equation \ref{eq:MAL} can be derived by starting with the integral of the age $x+t$ at each death child weighted by fertility and survival function

@@

## 5. Applications

The proposed meassures are applied to two different demographic regimes.

[PENDING]

A computation approach is used to evaluate the relative contribution of each component given a relative change in $\delta$.

[PENDING]

## 5. Bibliography
